### 2023-11-29 paper.tex

%%
%% Copyright 2022 OXFORD UNIVERSITY PRESS
%%
%% This file is part of the 'oup-authoring-template Bundle'.
%% ---------------------------------------------
%%
%% It may be distributed under the conditions of the LaTeX Project Public
%% License, either version 1.2 of this license or (at your option) any
%% later version.  The latest version of this license is in
%%    http://www.latex-project.org/lppl.txt
%% and version 1.2 or later is part of all distributions of LaTeX
%% version 1999/12/01 or later.
%%
%% The list of all files belonging to the 'oup-authoring-template Bundle' is
%% given in the file `manifest.txt'.
%%
%% Template article for OXFORD UNIVERSITY PRESS's document class `oup-authoring-template'
%% with bibliographic references
%%

%%%CONTEMPORARY%%%
%\documentclass[unnumsec,webpdf,contemporary,large]{oup-authoring-template}%
%\documentclass[unnumsec,webpdf,contemporary,large,namedate]{oup-authoring-template}% uncomment this line for author year citations and comment the above
%\documentclass[unnumsec,webpdf,contemporary,medium]{oup-authoring-template}
%\documentclass[unnumsec,webpdf,contemporary,small]{oup-authoring-template}

%%%MODERN%%%
%\documentclass[unnumsec,webpdf,modern,large]{oup-authoring-template}
\documentclass[unnumsec,webpdf,modern,large,namedate]{oup-authoring-template}% uncomment this line for author year citations and comment the above
%\documentclass[unnumsec,webpdf,modern,medium]{oup-authoring-template}
%\documentclass[unnumsec,webpdf,modern,small]{oup-authoring-template}

%%%TRADITIONAL%%%
%\documentclass[unnumsec,webpdf,traditional,large]{oup-authoring-template}
%\documentclass[unnumsec,webpdf,traditional,large,namedate]{oup-authoring-template}% uncomment this line for author year citations and comment the above
%\documentclass[unnumsec,namedate,webpdf,traditional,medium]{oup-authoring-template}
%\documentclass[namedate,webpdf,traditional,small]{oup-authoring-template}

%\onecolumn % for one column layouts

%\usepackage{showframe}
\usepackage{fontenc}

\usepackage{graphicx}

% line numbers
% \usepackage[mathlines, switch]{lineno}
% \usepackage[right]{lineno}

\begin{document}

% \journaltitle{Bioinformatics}
\journaltitle{Preprint}
% \DOI{DOI HERE}
\copyrightyear{2023}
% \pubyear{2019}
% \access{Advance Access Publication Date: Day Month Year}
\appnotes{Paper}

\firstpage{1}

%\subtitle{Subject Section}

\title[tstrait]{tstrait: a quantitative trait simulator for ancestral recombination graphs}

\author[1,2,$\ast$]{Daiki Tagami\ORCID{0000-0002-0923-1070}}
\author[2]{Gertjan Bisschop\ORCID{0000-0001-8327-0142}}
\author[2]{Jerome Kelleher\ORCID{0000-0002-7894-5253}}

\authormark{D.Tagami et al.}

\address[1]{\orgdiv{Department of Statistics}, \orgname{University of Oxford},
\orgaddress{\street{24-29 St Giles'}, \postcode{Oxford OX1 3LB},
\country{United Kingdom}}}
\address[2]{\orgdiv{Big Data Institute,
Li Ka Shing Centre for Health Information and Discovery},
\orgname{University of Oxford}, \orgaddress{\street{Old Road Campus},
\postcode{Oxford OX3 7LF}, \country{United Kingdom}}}

\corresp[$\ast$]{Corresponding author.
\href{email:daiki.tagami@hertford.ox.ac.uk}{daiki.tagami@hertford.ox.ac.uk}}

\received{Date}{0}{Year}
\revised{Date}{0}{Year}
\accepted{Date}{0}{Year}

\abstract{
\textbf{Summary:}
Ancestral recombination graphs (ARGs) encode the ensemble of correlated
genealogical trees arising from recombination in a compact and efficient
structure, and are of fundamental importance in population and statistical
genetics. Recent breakthroughs have made it possible to simulate and
infer ARGs at biobank scale, and there is now intense interest
in using ARG-based methods across a broad range of applications,
particularly in genome-wide association studies (GWAS).
Sophisticated methods exist to simulate ARGs using
population genetics models, but there is currently no software
to simulate quantitative traits directly from these ARGs.
To apply existing quantitative trait simulators users must export
genotype data, losing important information about ancestral processes
and producing prohibitively large files when applied
to the biobank-scale datasets currently of interest in GWAS.
We present \texttt{tstrait},
an open-source Python library to simulate
quantitative traits on ARGs, and show how this user-friendly software
can quickly simulate phenotypes for biobank-scale datasets
on a laptop computer.
\\\textbf{Availability and Implementation:} \texttt{tstrait} is available
for download on the Python Package Index.
Full documentation with examples and workflow
templates is available on \url{https://tskit.dev/tstrait/docs/}, and the
development version is maintained on GitHub
(\url{https://github.com/tskit-dev/tstrait}).\\ \textbf{Contact:}
\href{daiki.tagami@hertford.ox.ac.uk}{daiki.tagami@hertford.ox.ac.uk}\\
}

%\abstract{Abstracts must be able to stand alone and so cannot contain %citations to
%the paper's references, equations, etc. An abstract must consist of a single
%paragraph and be concise. Because of online formatting, abstracts must appear
%as plain as possible.}
\keywords{Ancestral Recombination Graph, ARG, GWAS, quantitative traits,
simulation}

\maketitle

\section{Introduction}

Genome-wide association studies (GWAS)
identify genetic variants that are statistically
associated with a specific trait~\citep{uffelmann2021}.
Many loci that are associated with various human diseases and
traits have been
identified~\citep[e.g.][]{yengo2022,mathieson2023},
% \citep{locke2015,ishigaki2022,
% mahajan2022,yengo2022,mathieson2023},
and GWAS results are actively being incorporated into clinical
practice~\citep{visscher2017}.
The great success of GWAS has prompted the collection of many biobank
datasets consisting of hundreds of thousands of
participants~\citep{tanjo2021practical},
but this scale presents significant challenges to
current GWAS methodology~\citep{uffelmann2021}.

Simulation is a critical component of GWAS method development, and
generally consists of two steps:
first simulating genetic variation (genotypes)
and then simulating quantitative traits (phenotypes) based on the
synthetic genotypes.
The combined genotypes and phenotypes represent ground-truth data
which GWAS methods can be evaluated against.
Genetic variation is
usually simulated either by model-based population genetic
methods such as
\texttt{msprime}~\citep{baumdicker2022efficient}
or \texttt{SLiM}~\citep{haller2023},
or by statistical resampling from existing datasets
using methods like \texttt{HAPGEN2}~\citep{su2011hapgen2}
or \texttt{HAPNEST}~\citep{wharrie2023hapnest}.
Both approaches have advantages and disadvantages and excel
in different situations.
Roughly speaking, model-based simulation methods
provide better control of population processes such as demography,
whereas resampling methods are better at capturing
difficult to model nuances of real data.
Model-based population genetic simulations have made
great strides in recent years, with major advances in both
scalability~\citep{kelleher2016efficient,kelleher2018efficient,haller2018tree}
and realism~\citep{adrion2020,anderson2023}, and
have been successfully used to simulate large-scale GWAS
cohorts~\cite[e.g.][]{martin2017,zaidi2020}.

An important property of these population genetic simulation methods is that
they output ancestral recombination graphs (ARGs)
rather than sample genotypes.
ARGs encode the interwoven paths
of genetic inheritance arising from
% Explanation of citations: people might get offended if we don't cite
% Griffiths when first mentioning the ARG, but Hudson has just as
% much to do with the actual definition that we use in Wong 2023.
recombination~\citep{hudson1983properties,griffiths1997ancestral,wong2023general},
and contain rich detail about ancestral processes.
Recent breakthroughs in inferrence methods have made it possible
to estimate ARGs at biobank scale~\citep{kelleher2019,zhang2023},
and there is now intense interest in their practical
application~\citep{lewanski2023era}.
Statistical genetics has been a particular focus,
and ARG-based methods have been shown
to detect more ultra rare variants than conventional association testing
methods~\citep{zhang2023};
have better power to detect causal loci in
quantitative-trait locus mapping~\citep{link2023tree};
and to provide a sparse and efficient model of linkage disequilibrium
in GWAS and downstream applications~\citep{nowbandegani2023extremely}.

ARG-based methods can simulate genetic variation for millions of
samples and store the output very compactly
using the ``succinct tree sequence'' encoding~\citep{wong2023general}.
and \texttt{tskit} library~\citep{ralph2020}.
For example, a highly realistic simulation
of chromosome 9 for 1.4 million
French-Canadian individuals~\citep{anderson2023}
requires around 550GB of storage space in gzip-compressed
VCF~\citep{danecek2011}.
At 1.36GB the original simulated ARG
(compressed using the \texttt{tszip} utility) is around 400X smaller.
Furthermore, many computations can
be expressed efficiently in terms of the underlying
ARG~\citep{kelleher2016efficient,ralph2020}, without
needing to decode the actual variation data.
Finally, outputting a simulated ARG provides access to the full
history, not just the genetic variation among the samples.

Although there are sophisticated methods available for simulating
ARGs at biobank scale, there is currently no
easy way to simulate quantitative traits based on such an ARG.
Many existing methods to simulate quantitative traits from a given set of
genetic sequences assume that the genotypes fit in
memory~\citep[e.g.][]{meyer2018,fernandes2020}, which makes them
impractical at biobank scale (the French-Canadian dataset discussed
above would require 140TB of RAM assuming 1 byte per genotype).
Methods that read parts of the genotype matrix from file as
required~\citep[e.g.][]{wharrie2023hapnest} can be used on
reasonable hardware, but working with such large files is slow
and cumbersome.
More fundamentally, exporting genotypes discards much of rich
detail about ancestral history contained in an ARG, and
it is exactly this information that we wish to take
advantage of when using inferred ARGs in GWAS applications.
In their analysis of the portability of polygenic risk
scores across populations, \citet{martin2017}
demonstrated the utility of simulating phenotypes directly
from an ARG. Their approach, however, is tightly coupled to the details
of the study and not designed to be reused.
Simulation code can be subtle and difficult to
debug~\citep{ragsdale2020lessons}, and there
is a critical need for a well-documented and
thoroughly tested means of simulating quantitative
traits directly from an ARG.

In this paper we present \texttt{tstrait}, a Python library that
efficiently simulates quantitative traits on an arbitrary ARG.
% in \texttt{tskit} format.
\texttt{Tstrait} can
quickly simulate quantitative traits for population-scale datasets,
with a very low memory overhead, and taking into account
the rich historical detail contained within an ARG.
The \texttt{tstrait} library also integrates well with the wider
Python data-science ecosystem~\citep{numpy},
allowing users to efficiently analyse
large-scale data using familiar and ergonomic tools.

\section{Results}
\begin{figure}[t]%
\centering
\includegraphics[width=240pt]{figures/tree-illustration.pdf}
\caption{Example simulation of a phenotype at a site with ancestral state A and
two mutations. In this diploid example each of the three individuals
is associated with two nodes (i.e., the individual with ID 0 corresponds
to nodes \textsf{a} and \textsf{b}). Internal nodes in the tree are associated
with the null individual, $-1$. Here, the trait's causal allele is T
and effect size $\beta=0.05$. Each node in the tree has an associated
genetic value, and the overall genetic value for an individual is the
sum of their corresponding node values. The final phenotype
for each individual is the sum of the genetic value and
simulated environmental noise.
\label{fig:tree-illustration}}
\end{figure}

\subsection{Model}
Phenotypes are simulated in \texttt{tstrait} following standard
GWAS models~\citep{uffelmann2021}, adapted to the ARG context.
Each trait is associated with one or more causal sites
(positions on the genome), and at each causal site there is a causal allele
(i.e.\ a particular nucleotide) associated with an effect size $\beta$.
For each causal site, an effect size is drawn from a distribution
and optionally multiplied by $\left(2p(1-p)\right)^\alpha$,
where $p$ is the frequency of the causal allele and $\alpha$
is a parameter describing the strength of
frequency dependence~\citep{speed2017}.
At a particular causal site, every node in the local tree that inherits
the causal allele at that site is said to have a ``genetic value'' of $\beta$.
% NOTE removing this for space - it keeps this section on one page.
% Because we are reasoning
% about the inheritance of alleles at a site on the genome,
% we can consider just the local tree at that site and
% the network structure of the ARG is not important~\citep{wong2023general}.

In Fig.~\ref{fig:tree-illustration} we show an example
tree for three individuals.
Because these individuals are diploids, each is associated
with two nodes in the tree (highlighted by colour).
Ancestral nodes are not associated with individuals here,
but in general an ARG may be embedded in a multigenerational
pedigree, where some internal nodes would be associated
with individuals.
In the example of
Fig.~\ref{fig:tree-illustration}, T is chosen as the causal
allele and $\beta=0.05$, and
therefore all nodes descending from \textsf{i}
have genetic value $0.05$, except \textsf{e} which
has zero because of the back-mutation to A.
Following the standard practise in GWAS~\citep{uffelmann2021},
we assume the additive model such that the overall
genetic value of an individual is the sum of its
nodes' genetic values.
Given these per-individual genetic values, the final phenotype
is then generated by adding some environmental noise.
This noise is simulated from a normal distribution with mean zero
and variance of $V_G(1-h^2)/{h^2}$,
where $V_G$ is the variance of the individual genetic values
and $h^2$ is the narrow-sense heritability provided as input by the user.

\subsection{Interface}
\texttt{Tstrait} is a Python library, building on the \texttt{tskit}
ARG toolkit~\citep{ralph2020,wong2023general} and the rich
Python data-science ecosystem~\citep{numpy}.
Simulating a phenotype for an ARG with default parameter
values requires only a few lines of code:
% NOTE this listing assumes num_causal has been moved, or removed.
% We need to check whether this actually works in the final version.
\begin{lstlisting}[language=Python,aboveskip=1em,belowskip=1em]
import tstrait as tst
model = tst.trait_model('normal', mean=0, var=1)
result = tst.sim_phenotype(arg, model)
\end{lstlisting}
We first create \texttt{model}, which represents the distribution
from which effect sizes are drawn. Five commonly used
univariate distributions are supported, along with the
multivariate normal distribution to model pleiotropic traits.
Given this model, we can then simulate phenotypes for the individuals
in an ARG (as a  \texttt{tskit TreeSequence}) using the
\texttt{sim\_phenotype} function.
The user can either specify a number of causal sites to be chosen randomly
along the genome (one, by default), or can directly provide the causal
sites as input. Combined with the detailed information about
mutations recorded in a \texttt{tskit} ARG, explicitly specifying causal
sites allows us to model many different types of trait, for example
those associated with mutations arising in a particular population
or time interval.
The return value \texttt{result} is an object encapsulating
two Pandas dataframes~\citep{mckinney2010data}: one describing the simulated
effect sizes and the other describing the genetic values,
environmental noise and final phenotypes for each individual.
The simulation results can then be efficiently and conveniently
processed using familiar Python data science tools.

As well as this convenient single-function interface,
\texttt{tstrait} provides modular building blocks for power-users
and to facilitate integration with other tools that generate
traits on an ARG. The \texttt{sim\_trait} function
simulates effect sizes for an input ARG, and returns a data frame
describing the causal sites, alleles and effect sizes.
This dataframe can then be passed to the \texttt{genetic\_values}
function, which calculates the genetic values for each
node, and accumulates them by individual
(Fig~\ref{fig:tree-illustration}). Finally, the \texttt{sim\_env}
function takes these per-individual genetic values
and adds some simulated environmental noise to produce the final
phenotypes.

A major benefit of this modular architecture is the flexibility it offers
users. Because the causal sites and effect sizes are specified in a
simple tabular format, users can easily develop their own approach
to simulating these values. Alternatively, other simulators
such as SLiM~\citep{haller2023} that generate effect sizes and
causal mutations during the progress of a forwards-time simulation
could output these
values to a CSV or similar file. The modular architecture  and simple
input data formats are specifically intended to facilitate such
interoperability.

\subsection{Implementation and validation}
\texttt{Tstrait} is written entirely in Python. Numerical operations
are either peformed using standard array-oriented
approaches~\citep{numpy} or accelerated
using the \texttt{numba} JIT compiler~\citep{numba}.
The \texttt{tstrait} codebase includes an extensive suite of unit tests,
which are automatically run as part of the development process. The
output of \texttt{tstrait} has been extensively validated against
theoretical expectations, as well as the output
of \texttt{AlphaSimR}~\citep{gaynor2021} and
\texttt{simplePHENOTYPES}~\citep{fernandes2020}.

\subsection{Performance}
\texttt{Tstrait} is very efficient, and can be applied to datasets at the
largest scales on standard computers.
Supplementary Fig~\ref{fig:time} shows how trait simulation time scales
with number of individuals on human-like coalescent simulations
generated using~\texttt{stdpopsim}~\citep{adrion2020}.
For each ARG, we simulate a trait with 1000 causal sites
and report the mean CPU (Intel(R) Core(TM) i9-11900H)
time over 10 independent replicates.
To emphasise scalability, we also
applied \texttt{tstrait} to the simulations of 1.4 million
French-Canadians discussed in the introduction.
% TODO redo this for 100 causal sites @daiki.
It took 6 seconds to simulate a trait with 100 causal sites.
% Is this worth the space? We should cite 1KG project also really.
Finally, to demonstrate that \texttt{tstrait} can also be applied
to ARGs inferred from real data, we simulated a trait with 100 causal
sites for an ARG estimated from 1000 Genomes project
data~\citep{kelleher2019} which has YYY samples
and ZZZ variant sites. This took 0.X seconds.
Memory requirements for \texttt{tstrait} are modest: all of the
above experiments were performed on a laptop computer with 16GB of RAM.

\section{Conclusion}
There is substantial interest in using inferred ARGs to improve
association testing
methods~\citep{zhang2023,link2023tree,nowbandegani2023extremely},
and there is a pressing need for a well-tested, efficient
and user-friendly means of simulating phenotypes on ARGs.
Highly realistic simulations conditioned on large
pedigrees~\citep{anderson2023} provide an exciting opportunity
to test the effects of intricate population structure on
GWAS, and we hope that \texttt{tstrait} will facilitate these
investigations. \texttt{Tstrait}'s modular architecture
and flexible specification of causal sites
should provide the opportunity to explore new avenues of research,
and an extensible platform for future development.

\section{Competing interests}
No competing interest is declared.

\section{Acknowledgments}
We are grateful to Gregor Gorjanc, Ben Haller, Ben Jeffery and
others in the tskit community for helpful discussions and feedback.

\section{Funding}
DT is supported by the Oxford Kobe Scholarship from the University of Oxford
and the Euretta J. Kellett Fellowship from Columbia University.
JK acknowledges support from the Robertson Foundation,
NIH (research grants HG011395 and HG012473) and
EPSRC (research grant EP/X024881/1).

%USE THE BELOW OPTIONS IN CASE YOU NEED AUTHOR YEAR FORMAT.
\bibliographystyle{abbrvnat}
\bibliography{paper}

\clearpage

\renewcommand\thefigure{S\arabic{figure}}
\setcounter{figure}{0}
\renewcommand\thetable{S\arabic{table}}
\setcounter{table}{0}
\section{Supplementary Figures}

\begin{figure*}[t]%
\centering
% FIXME get the sizes right here and in the notebook where we're generating
% the figure so we don't have to scale here.
\includegraphics[width=213pt]{figures/time-scaling.pdf}
\caption{\textbf{Time taken to simulate quantitative traits.} Each point
represents the mean time for 10 independent runs under different random seeds.
The times reported are the total CPU time required to simulate quantitative
traits on an Intel(R) Core(TM) i9-11900H CPU and 16 GB of RAM. The trait model
is a normal distribution with $\mu=0$, $\sigma^2=1$, $h^2=0.3$, and
$\alpha=-1$. (A) CPU time with increasing number of individuals
based on ARGs simulated using \texttt{stdpopsim} with the [MODEL ID]
demographic mode. For each replicate we simulate
one quantitative trait with 1000 causal sites.
(B) CPU time with increasing number of causal sites on two large ARGs.
\texttt{tstrait} is used to simulate quantitative traits from
the simulated French Canadian dataset \citep{anderson2023} and the inferred
tree sequence dataset of the 1000 Genomes Project \citep{kelleher2019} with
varying number of causal sites. The French Canadian dataset is downloaded from
\url{https://zenodo.org/record/6839683}, and the inferred ARG from the 1000
Genomes Project is downloaded from \url{https://zenodo.org/record/3051855}.
Chromosome 9 is selected for both datasets.}\label{fig:time}
\end{figure*}

\end{document}

### 2023-12-04 paper.tex
Comments from Daiki

@@
 \texttt{msprime}~\citep{baumdicker2022efficient}
-or \texttt{SLiM}~\citep{haller2023},
+and \texttt{SLiM}~\citep{haller2023},
 or by statistical resampling from existing datasets
 using methods like \texttt{HAPGEN2}~\citep{su2011hapgen2}
-or \texttt{HAPNEST}~\citep{wharrie2023hapnest}.
+and \texttt{HAPNEST}~\citep{wharrie2023hapnest}.
 Both approaches have advantages and disadvantages and excel
@@
 to detect more ultra rare variants than conventional association testing
-methods~\citep{zhang2023};
+methods~\citep{zhang2023}; to
 have better power to detect causal loci in
@@
 samples and store the output very compactly
-using the ``succinct tree sequence'' encoding~\citep{wong2023general}.
+in the ``succinct tree sequence'' encoding~\citep{wong2023general}
 and \texttt{tskit} library~\citep{ralph2020}.
@@
 (compressed using the \texttt{tszip} utility) is around 400X smaller.
-Furthermore, many computations can
+Furthermore, many calculations can
 be expressed efficiently in terms of the underlying
@@
 with the null individual, $-1$. Here, the trait's causal allele is T
-and effect size $\beta=0.05$. Each node in the tree has an associated
+with an effect size $\beta=0.05$. Each node in the tree has an associated
 genetic value, and the overall genetic value for an individual is the
-sum of their corresponding node values. The final phenotype
+sum of the genetic values of their corresponding nodes. 
+The final phenotype
 for each individual is the sum of the genetic value and
@@
 Fig.~\ref{fig:tree-illustration}, T is chosen as the causal
-allele and $\beta=0.05$, and
-therefore all nodes descending from \textsf{i}
+allele with $\beta=0.05$, 
+so all nodes descending from \textsf{i}
 have genetic value $0.05$, except \textsf{e} which
@@
 effect sizes and the other describing the genetic values,
-environmental noise and final phenotypes for each individual.
+environmental noise and phenotypes for each individual.
 The simulation results can then be efficiently and conveniently

### 2024-03-11 paper.tex
Minor changes from Daiki

@@
 For each causal site, an effect size is drawn from a distribution
-and optionally multiplied by $\left(2p(1-p)\right)^\alpha$,
+and optionally multiplied by $\left(2p(1-p)\right)^{\alpha/2}$,
 where $p$ is the frequency of the causal allele and $\alpha$
 is a parameter describing the strength of
-frequency dependence~\citep{speed2017}.
+frequency dependence~\citep{speed2012}.
 At a particular causal site, every node in the local tree that inherits
@@
 To emphasise scalability, we also
-applied \texttt{tstrait} to the simulations of 1.4 million
+% TODO check that the numbers align here, we may need to be 
+% less specific to avoid confusing readers or having to
+% explain technical details.
+applied \texttt{tstrait} to the simulations of 2.7 million
 French-Canadians discussed in the introduction.
+% tstrait simulates phenotypes for 2.7 million individuals in the French Canadian dataset, as ancestors are denoted as individuals in the dataset.
 % TODO redo this for 100 causal sites @daiki.
-It took 6 seconds to simulate a trait with 100 causal sites.
+It took 80.69 seconds to simulate a trait with 100 causal sites.
 % Is this worth the space? We should cite 1KG project also really.
@@
 sites for an ARG estimated from 1000 Genomes project
-data~\citep{kelleher2019} which has YYY samples
-and ZZZ variant sites. This took 0.X seconds.
+data~\citep{kelleher2019} which has 2,504 samples
+and 1,685,401 variant sites. This took 5.40 seconds.
 Memory requirements for \texttt{tstrait} are modest: all of the
@@
 $\alpha=-1$. (A) CPU time with increasing number of individuals
-based on ARGs simulated using \texttt{stdpopsim} with the [MODEL ID]
+based on ARGs simulated using \texttt{stdpopsim} with the HomSap
 demographic mode. For each replicate we simulate
@@
 
+\begin{figure*}[t]%
+\centering
+% FIXME get the sizes right here and in the notebook where we're generating
+% the figure so we don't have to scale here.
+\includegraphics[width=213pt]{figures/time-scaling.pdf}
+\caption{\textbf{Time taken to simulate quantitative traits.} Each point
+represents the mean time for 10 independent runs under different random seeds with 95\% confidence intervals plotted.
+The times reported are the total CPU time required to simulate quantitative
+traits on an Intel(R) Core(TM) i9-11900H CPU and 16 GB of RAM. The trait model
+is a normal distribution with $\mu=0$, $\sigma^2=1$, $h^2=0.3$, and
+$\alpha=0$.
+\texttt{tstrait} is used to simulate quantitative traits from ARGs simulated using \texttt{stdpopsim} with the HomSap demographic mode.
+For each replicate we simulate one quantitative trait with 1000 causal sites.
+We have compared \texttt{tstrait} against the quantitative trait simulation framework that is based on \texttt{tskit} functionalities.
+The comparison codes are adapted from \citep{martin2017}, and they are modified based on functionalities offered in \texttt{tskit} version 0.5.6 to speed up the computational efficiency.}\label{fig:time}
+\end{figure*}
 \end{document}

### 2024-03-11 paper.tex
Updates from Overleaf

@@
 $\alpha=0$.
-\texttt{tstrait} is used to simulate quantitative traits from ARGs simulated using \texttt{stdpopsim} with the HomSap demographic mode.
+\texttt{tstrait} is used to simulate quantitative traits from ARGs simulated using \texttt{stdpopsim} with the generic piecewise constant size model in Homo Sapiens catalog.
 For each replicate we simulate one quantitative trait with 1000 causal sites.

### 2024-03-12 paper.tex
Daiki's changes

@@
 % the figure so we don't have to scale here.
-\includegraphics[width=213pt]{figures/time-scaling.pdf}
-\caption{\textbf{Time taken to simulate quantitative traits.} Each point
-represents the mean time for 10 independent runs under different random seeds.
-The times reported are the total CPU time required to simulate quantitative
-traits on an Intel(R) Core(TM) i9-11900H CPU and 16 GB of RAM. The trait model
-is a normal distribution with $\mu=0$, $\sigma^2=1$, $h^2=0.3$, and
-$\alpha=-1$. (A) CPU time with increasing number of individuals
-based on ARGs simulated using \texttt{stdpopsim} with the HomSap
-demographic mode. For each replicate we simulate
-one quantitative trait with 1000 causal sites.
-(B) CPU time with increasing number of causal sites on two large ARGs.
-\texttt{tstrait} is used to simulate quantitative traits from
-the simulated French Canadian dataset \citep{anderson2023} and the inferred
-tree sequence dataset of the 1000 Genomes Project \citep{kelleher2019} with
-varying number of causal sites. The French Canadian dataset is downloaded from
-\url{https://zenodo.org/record/6839683}, and the inferred ARG from the 1000
-Genomes Project is downloaded from \url{https://zenodo.org/record/3051855}.
-Chromosome 9 is selected for both datasets.}\label{fig:time}
-\end{figure*}
-
-\begin{figure*}[t]%
-\centering
-% FIXME get the sizes right here and in the notebook where we're generating
-% the figure so we don't have to scale here.
-\includegraphics[width=213pt]{figures/time-scaling.pdf}
+\includegraphics[width=\columnwidth]{figures/computational-time-code/stdpopsim-time.pdf}
 \caption{\textbf{Time taken to simulate quantitative traits.} Each point

### 2024-03-12 paper.tex

@@
 of chromosome 9 for 1.4 million
-French-Canadian individuals~\citep{anderson2023}
+French-Canadian samples~\citep{anderson2023}
 requires around 550GB of storage space in gzip-compressed
@@
 using the \texttt{numba} JIT compiler~\citep{numba}.
-The \texttt{tstrait} codebase includes an extensive suite of unit tests,
+The \texttt{tstrait} codebase includes a suite of unit tests,
 which are automatically run as part of the development process. The
-output of \texttt{tstrait} has been extensively validated against
+output of \texttt{tstrait} has been validated against
 theoretical expectations, as well as the output
@@
 generated using~\texttt{stdpopsim}~\citep{adrion2020}.
-For each ARG, we simulate a trait with 1000 causal sites
-and report the mean CPU (Intel(R) Core(TM) i9-11900H)
-time over 10 independent replicates.
 To emphasise scalability, we also
@@
 % explain technical details.
-applied \texttt{tstrait} to the simulations of 2.7 million
-French-Canadians discussed in the introduction.
-% tstrait simulates phenotypes for 2.7 million individuals in the French Canadian dataset, as ancestors are denoted as individuals in the dataset.
-% TODO redo this for 100 causal sites @daiki.
-It took 80.69 seconds to simulate a trait with 100 causal sites.
+applied \texttt{tstrait} to the large
+French-Canadian simulations discussed in the introduction.
+It took 80.69 seconds to simulate a trait with 100 causal sites
+for all 2.7 million pedigree individuals.
 % Is this worth the space? We should cite 1KG project also really.
@@
 
-\begin{figure*}[t]%
+\begin{figure}[t]%
 \centering
@@
 \includegraphics[width=\columnwidth]{figures/computational-time-code/stdpopsim-time.pdf}
-\caption{\textbf{Time taken to simulate quantitative traits.} Each point
-represents the mean time for 10 independent runs under different random seeds with 95\% confidence intervals plotted.
-The times reported are the total CPU time required to simulate quantitative
-traits on an Intel(R) Core(TM) i9-11900H CPU and 16 GB of RAM. The trait model
-is a normal distribution with $\mu=0$, $\sigma^2=1$, $h^2=0.3$, and
-$\alpha=0$.
-\texttt{tstrait} is used to simulate quantitative traits from ARGs simulated using \texttt{stdpopsim} with the generic piecewise constant size model in Homo Sapiens catalog.
-For each replicate we simulate one quantitative trait with 1000 causal sites.
-We have compared \texttt{tstrait} against the quantitative trait simulation framework that is based on \texttt{tskit} functionalities.
-The comparison codes are adapted from \citep{martin2017}, and they are modified based on functionalities offered in \texttt{tskit} version 0.5.6 to speed up the computational efficiency.}\label{fig:time}
-\end{figure*}
+\caption{Time taken to simulate quantitative traits with increasing
+sample size. For each sample size, we simulated an ARG under human-like
+parameters using the default HomSap demographic model in \texttt{stdpopsim}.
+Each point represents the mean time for 10 independent runs of \texttt{tstrait}
+for a particular ARG. The times reported are the total CPU
+time required to simulate a quantitative trait with 1000 causal sites,
+on an Intel(R) Core(TM) i9-11900H CPU and 16 GB of RAM.
+The trait model is a normal distribution with $\mu=0$,
+$\sigma^2=1$, $h^2=0.3$, and $\alpha=0$.
+}\label{fig:time}
+\end{figure}
 \end{document}

### 2024-03-12 cover_letter/cover_letter.tex
Update cover letter

\documentclass{letter}

\signature{Daiki Tagami}

\address{Big Data Institute\\University of Oxford\\UK}
\begin{document}

\begin{letter}{Bioinformatics}

\opening{Dear Editors,}

I am writing on behalf of my coauthors to submit our manuscript entitled
\emph{tstrait: a quantitative trait simulator for ancestral recombination graphs}
for consideration for publication in Bioinformatics
as an ``Application Note'' research article.

Following recent breakthroughs in simulation and inference,
Ancestral Recombination Graphs (ARGs) are the subject of
intense research interest (see Lewanski et al 2024, PLoS Genetics,
and Brandt et al 2024, GBE, for recent perspectives). One of the key application
areas is statistical genetics, and in particular developing
improved methodology for Genome Wide Association Studies (GWAS).
However, there is currently no available software to efficiently
and conveniently simulate quantitative traits from an ARG,
significantly hampering progress. In this manuscript we describe
\texttt{tstrait}, a new Python package that can efficiently
simulate quantitative traits on ARGs under a wide range of
models. We believe that this tool will be of significant interest
to the readers of Bioinformatics, and will enable further
ground-breaking research in this area.

\closing{Sincerely,}

\end{letter}
\end{document}

### 2024-03-12 paper.tex

@@
 and there is now intense interest in their practical
-application~\citep{lewanski2023era}.
+application~\citep{lewanski2024era,brandt2024promise}.
 Statistical genetics has been a particular focus,

### 2024-03-13 paper.tex
Minor wording tweaks

@@
 % \DOI{DOI HERE}
-\copyrightyear{2023}
+\copyrightyear{2024}
 % \pubyear{2019}
@@
 ARGs encode the interwoven paths
-of genetic inheritance arising from
+of genetic inheritance caused by
 % Explanation of citations: people might get offended if we don't cite
@@
 % much to do with the actual definition that we use in Wong 2023.
-recombination~\citep{hudson1983properties,griffiths1997ancestral,wong2023general},
+recombination \citep{hudson1983properties,griffiths1997ancestral,wong2023general},
 and contain rich detail about ancestral processes.
@@
 \end{lstlisting}
-We first create \texttt{model}, which represents the distribution
+We first create \texttt{model}, representing the distribution
 from which effect sizes are drawn. Five commonly used
@@
 The simulation results can then be efficiently and conveniently
-processed using familiar Python data science tools.
+processed using standard Python data science tools.
 

### 2024-03-13 figures/computational-time-code/tskit_phenotype.py
remove unused files

@@ -1,54 +1,0 @@
-# This simulation code is adapted from the true_prs function in https://github.com/armartin/ancestry_pipeline/blob/master/simulate_prs.py#L111
-# The simulation framework is described in Martin et al. (2017), https://www.sciencedirect.com/science/article/pii/S0002929717301076
-import numpy as np
-from dataclasses import dataclass
-import tskit
-
-@dataclass
-class Result:
-    phenotype: np.ndarray
-    site_id: np.ndarray
-    causal_allele: np.ndarray
-    effect_size: np.ndarray
-    allele_frequency: np.ndarray    
-
-
-def sim_phenotype_tskit(ts, h2, ncausal, random_seed):
-    
-    rng = np.random.default_rng(random_seed)
-    
-    causal_site_index = np.linspace(0, ts.num_sites-1, ncausal, dtype=int)
-    
-    allele_freq = np.zeros(ncausal, dtype=float)
-    effect_size = np.zeros(ncausal, dtype=float)
-    causal_allele = np.zeros(ncausal, dtype=object)
-    
-    nhaps = ts.num_samples
-
-    prs_haps = np.zeros(nhaps) #score for each haplotype
-    i = 0
-    variant = tskit.Variant(ts)
-    for causal_site in causal_site_index:
-        variant.decode(causal_site)
-        sim_effect_size = np.random.normal(loc=0,scale=h2/ncausal)
-        effect_size[i] = sim_effect_size
-        causal_allele_index = rng.integers(low=1, high=len(variant.alleles))
-        allele = variant.alleles[causal_allele_index]
-        causal_allele[i] = allele
-        genotype = np.array([1 if i == causal_allele_index else 0 for i in variant.genotypes])
-        prs_haps += genotype * sim_effect_size
-        allele_freq[i] = np.sum(genotype) / len(genotype)
-        i += 1
-
-    env_effect = rng.normal(loc=0,scale=1-h2, size=nhaps)
-    prs_norm = (prs_haps - np.mean(prs_haps)) / np.std(prs_haps)
-    env_norm = (env_effect - np.mean(env_effect)) / np.std(env_effect)
-    total_liability = np.sqrt(h2) * prs_norm + np.sqrt(1 - h2) * env_norm
-    
-    simulation_result = Result(phenotype = total_liability,
-                                        site_id = causal_site_index,
-                                       causal_allele = causal_allele,
-                                       effect_size = effect_size,
-                                       allele_frequency = allele_freq)
-    
-    return simulation_result

### 2024-03-13 paper.tex
Fix figure
Updates from Overleaf

@@
 \section{Acknowledgments}
-We are grateful to Gregor Gorjanc, Ben Haller, Ben Jeffery and
+We are grateful to Gregor Gorjanc, Ben Haller, Ben Jeffery, Pier Palamara, Alison Etheridge and
 others in the tskit community for helpful discussions and feedback.
@@
 
-\clearpage
+%\clearpage
 
@@
 
-\begin{figure}[t]%
+\begin{figure}[h]%
 \centering
@@
 % the figure so we don't have to scale here.
-\includegraphics[width=\columnwidth]{figures/computational-time-code/stdpopsim-time.pdf}
+\includegraphics[width=\columnwidth]{figures/stdpopsim-time.pdf}
 \caption{Time taken to simulate quantitative traits with increasing

### 2024-05-13 response-to-reviewer/response-to-reviewer.tex
Revision

% LaTeX rebuttal letter example.
%
% Copyright 2019 Friedemann Zenke, fzenke.net
%
% Based on examples by Dirk Eddelbuettel, Fran and others from
% https://tex.stackexchange.com/questions/2317/latex-style-or-macro-for-detailed-response-to-referee-report
%
% Licensed under cc by-sa 3.0 with attribution required.

\documentclass[11pt]{article}
\usepackage[utf8]{inputenc}
\usepackage{lipsum} % to generate some filler text
\usepackage{fullpage}

% import Eq and Section references from the main manuscript where needed
% \usepackage{xr}
% \externaldocument{manuscript}

% package needed for optional arguments
\usepackage{xifthen}
% define counters for reviewers and their points
\newcounter{reviewer}
\setcounter{reviewer}{0}
\newcounter{point}[reviewer]
\setcounter{point}{0}

% This refines the format of how the reviewer/point reference will appear.
\renewcommand{\thepoint}{\thereviewer.\arabic{point}}

% command declarations for reviewer points and our responses
\newcommand{\reviewersection}{\stepcounter{reviewer} \bigskip \hrule
                  \section*{Reviewer \thereviewer}}

\newenvironment{point}
   {\refstepcounter{point} \bigskip \noindent {\textbf{Reviewer~Point~\thepoint} } ---\ }
   {\par }

\newcommand{\shortpoint}[1]{\refstepcounter{point}  \bigskip \noindent
    {\textbf{Reviewer~Point~\thepoint} } ---~#1\par }

\newenvironment{reply}
   {\medskip \noindent \begin{sf}\textbf{Reply}:\  }
   {\medskip \end{sf}}

\newcommand{\shortreply}[2][]{\medskip \noindent \begin{sf}\textbf{Reply}:\  #2
    \ifthenelse{\equal{#1}{}}{}{ \hfill \footnotesize (#1)}%
    \medskip \end{sf}}

\begin{document}

\section*{Response to the editor}
% General intro text goes here
Thank you for considering this manuscript for publication, and we are extremely excited to hear that this manuscript is acceptable for publication in Bioinformatics subject to the suggested revisions. We are glad that
Reviewers 1 and 2 share our enthusiasm for the tstrait package!
\subsection*{Associate Editor}
% We address your specific comments below:
% % \reviewersection

\textit{
The reviewers are both very positive on your manuscript, rating it highly and seeing no significant problems in need of revision.  Each has a couple of questions for you, and although the reviewers may consider these discretionary I will pass them on to you with a request for a Minor Revision to give you a chance to revise in response to these points before we make a final decision. Given the reviews, I think it is unlikely we will need a second round of review if you can provide reasonable answers to the questions raised.}

\begin{reply}
    Thank you for the helpful feedback. We are grateful for the reviewers 
    comments and have updated the manuscript to clarify.
\end{reply}

\section*{Response to the reviewers}
% General intro text goes here
We thank the reviewers for their close reading of our manuscript and
insightful comments. In the following we address the points raised
in turn.

% Let's start point-by-point with Reviewer 1
\reviewersection

% Point one description
\begin{point}
This applications note describes tstrait, which uses ARGs as a basis for simulating traits. This is important because simulated genotypic data use ARGs to quantify what is happening in the genome across many generations.

I thought this paper was very well-written. In particular, Figure 1 is very effective at presenting how traits are simulated. I also thought that the motivation behind the need for an ARG-based trait simulation package was strongly justified. If/when ARGs become more widely used, I think it will be important to have simulation packages like this.
\end{point}

\begin{reply}
    Thank you for the summary. We are delighted to hear your positive feedback.
\end{reply}

\begin{point}
    1) Can dominance and/or epistatic effects be simulated under the framework of, e.g., Figure 1?
\end{point}

\begin{reply}
    Both dominance and epistasis can be simulated under the simulation framework that is described in Figure 1, as it is possible to extract the genetic value of each node at every site. We have included this in the revised manuscript, and we are planning to incorporate these features in the future release.
\end{reply}

\begin{point}
    2) I appreciate the effort to put in the modular framework. With this said, how easy is it to take the simulated traits from tstrait and migrate the files over to R to run, e.g., GWAS and/or genomic prediction?
\end{point}

\begin{reply}
    All simulation results in tstrait are provided as pandas dataframe. Users can simply save the pandas dataframe as a CSV or similar file and load it into other programs. We have added this part in the revised manuscript to emphasize the point.
\end{reply}

\reviewersection

% Point one description
\begin{point}
Tagami et al. present an interesting Python library that can efficiently simulate quantitative traits from ancestral recombination graphs. I like how the authors illustrate the efficiency of this library when considering the size of typical biobank data.
\end{point}

\begin{reply}
    Thank you for your kind comments. We are delighted to hear your positive feedback on this new Python library.
\end{reply}

\begin{point}
    - I was unable to locate the comparisons against AlphaSimR and simplePHENOTYPES.
\end{point}

\begin{reply}
    The comparisons are located in verification.py file in the tstrait GitHub repository 
    to ensure that these statistical tests are done as part of the normal development 
    of the package. As a result, however, they are not very easy to find. We have 
    added a short supplemental note describing these validations, and describing
    where they can be found in the tstrait source code.
\end{reply}

\begin{point}
    Although I appreciate the argument of efficiency, it seems a little exaggerated since for some of those packages (e.g. simplePHENOTYPES), the simulation step only utilizes the QTNs. I.e., although the whole file would have 140 TB, this file would not be needed in its entirety. For a trait with 100 QTNs, only 100 SNPs would be needed for the simulation step.
\end{point}

\begin{reply}
    Although it is correct that methods such as simplePHENOTYPES only use a subset of 
    the genotype matrix, they must still read these large VCFs, which is itself a 
    major undertaking. In practise they read entire VCF into memory.
    For example, simplePHENOTYPES uses the SNPRelate package to first 
    convert the input VCF to GDS format, and then use the \texttt{snpgdsGetGeno}
    function to obtain an explicit genotype matrix in memory for the entire 
    dataset. The sites of interest are then extracted from this large matrix 
    for further analysis. While this could be done more efficiently, 
    any operations on a multi-terabyte VCF are cumbersome and would require
    substantial engineering effort.
\end{reply}

\begin{point}
    Another positive aspect is that the documentation on GitHub is well documented.
\end{point}

\begin{reply}
    Thank you for your feedback, and we agree that documentation is very important.
\end{reply}

\end{document}

### 2024-05-13 cover_letter/cover_letter_resubmission.tex
Revision

--- 2024-03-12 cover_letter/cover_letter.tex
@@
 
-I am writing on behalf of my coauthors to submit our manuscript entitled
+I am writing on behalf of my coauthors to resubmit our manuscript entitled
 \emph{tstrait: a quantitative trait simulator for ancestral recombination graphs}
@@
 
-Following recent breakthroughs in simulation and inference,
-Ancestral Recombination Graphs (ARGs) are the subject of
-intense research interest (see Lewanski et al 2024, PLoS Genetics,
-and Brandt et al 2024, GBE, for recent perspectives). One of the key application
-areas is statistical genetics, and in particular developing
-improved methodology for Genome Wide Association Studies (GWAS).
-However, there is currently no available software to efficiently
-and conveniently simulate quantitative traits from an ARG,
-significantly hampering progress. In this manuscript we describe
-\texttt{tstrait}, a new Python package that can efficiently
-simulate quantitative traits on ARGs under a wide range of
-models. We believe that this tool will be of significant interest
-to the readers of Bioinformatics, and will enable further
-ground-breaking research in this area.
+We thank the reviewers for their thoughtful and thorough reviews, and we have addressed those points in the revised manuscript. We have attached
+a detailed point-by-point response in the \texttt{response-to-reviewers.pdf}
+file.
+
+We have also attached a file showing the changes made to the manuscript
+in response to reviewers in \texttt{review-diff.pdf}.
+
+Finally, we attach \texttt{supplementary-material.pdf}, which includes some clarifications on our 
+statistical testing methodology.
 

### 2024-05-13 paper.tex
Revision

@@
 and $h^2$ is the narrow-sense heritability provided as input by the user.
+Dominance and epistasis are straightforward extensions to this
+model, which we plan to include in future versions of \texttt{tstrait}.
 
@@
 The simulation results can then be efficiently and conveniently
-processed using standard Python data science tools.
+processed using standard Python data science tools, or exported
+to (e.g.) CSV for broad compatibility.
 
@@
 
-\renewcommand\thefigure{S\arabic{figure}}
-\setcounter{figure}{0}
-\renewcommand\thetable{S\arabic{table}}
-\setcounter{table}{0}
-\section{Supplementary Figures}
-
-\begin{figure}[h]%
-\centering
-% FIXME get the sizes right here and in the notebook where we're generating
-% the figure so we don't have to scale here.
-\includegraphics[width=\columnwidth]{figures/stdpopsim-time.pdf}
-\caption{Time taken to simulate quantitative traits with increasing
-sample size. For each sample size, we simulated an ARG under human-like
-parameters using the default HomSap demographic model in \texttt{stdpopsim}.
-Each point represents the mean time for 10 independent runs of \texttt{tstrait}
-for a particular ARG. The times reported are the total CPU
-time required to simulate a quantitative trait with 1000 causal sites,
-on an Intel(R) Core(TM) i9-11900H CPU and 16 GB of RAM.
-The trait model is a normal distribution with $\mu=0$,
-$\sigma^2=1$, $h^2=0.3$, and $\alpha=0$.
-}\label{fig:time}
-\end{figure}
 \end{document}

### 2024-05-13 supplementary-material/supplementary-material.tex
Revision

\documentclass{article}
\usepackage{natbib}

\usepackage{graphicx}
\usepackage{url}

\title{Supplementary Material for tstrait: a quantitative trait simulator for ancestral recombination graphs}
\date{}
\usepackage[margin=1in]{geometry}

\begin{document}
\maketitle

\section{Supplementary Notes}

We present an overview of statistical tests that are conducted during the development process of the \texttt{tstrait} package. Three types of tests are conducted to validate the statistical properties of \texttt{tstrait}: exact tests, comparison tests, and statistical tests.

\subsection{Exact Tests}

We simulate effect sizes and phenotypes without environmental noise by using two different quantitative trait simulators, \texttt{AlphaSimR}~\citep{gaynor2021} and
\texttt{simplePHENOTYPES}~\citep{fernandes2020}, and the simulation framework described in \citep{zhang2023}. The simulated effect sizes are directly used in \texttt{tstrait} to simulate phenotypes. We show that when traits are influenced with the same causal sites and effect sizes, the simulated genetic values from \texttt{tstrait} are identical to the other simulation frameworks.

\subsection{Comparison Tests}

We simulate phenotypes with environmental noise in \texttt{AlphaSimR}~\citep{gaynor2021},
\texttt{simplePHENOTYPES}~\citep{fernandes2020} and the simulation framework described in \citep{zhang2023} by using the same parameters that are used in the \texttt{tstrait} simulation. By using QQ-plots, we show that the simulated traits from \texttt{tstrait} are having similar distributions as the traits simulated from other programs.

\subsection{Statistical Tests}

Environmental noise and effect sizes are simulated by using various parameters in \texttt{tstrait} and their statistical properties are validated by using QQ-pplots.

\subsection{Availability of Materials}

Code to run the statistical tests is included in the \texttt{tstrait} GitHub repository, \url{https://github.com/tskit-dev/tstrait/blob/main/verification.py}. These tests are used as part of the development process of the \texttt{tstrait} package.

\section{Supplementary Figures}

\renewcommand\thefigure{S\arabic{figure}}
\setcounter{figure}{0}
\renewcommand\thetable{S\arabic{table}}
\setcounter{table}{0}

\begin{figure}[h]%
\centering
\includegraphics[width=0.7\columnwidth]{figures/stdpopsim-time.pdf}
\caption{Time taken to simulate quantitative traits with increasing
sample size. For each sample size, we simulated an ARG under human-like
parameters using the default HomSap demographic model in \texttt{stdpopsim}.
Each point represents the mean time for 10 independent runs of \texttt{tstrait}
for a particular ARG. The times reported are the total CPU
time required to simulate a quantitative trait with 1000 causal sites,
on an Intel(R) Core(TM) i9-11900H CPU and 16 GB of RAM.
The trait model is a normal distribution with $\mu=0$,
$\sigma^2=1$, $h^2=0.3$, and $\alpha=0$.
}\label{fig:time}
\end{figure}

\newpage

\bibliographystyle{abbrvnat}
\bibliography{paper}

\end{document}
